Estimation of Intracortical Connectivity in a Dynamical Neural Field Model

### 2010-09-18 ltx/first_order_framework/LaTeX_Manuscript.tex
tried to narrow claims

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Since the mid 1970s there has been significant developments in generating neural field models that provide theoretical insights into macroscopic neurodynamics, which have previously been impossible to gain from experimental data alone. In parallel to these developments, technological advances have led to significant increases into the spatiotemporal resolution of data acquired in electrophysiological recording systems. In order to make the best of these developments, a method is required for relating data to the mathematical models.

This paper presents a framework for creating neural field models from patient-specific or experimental electrophysiological data. The Wilson and Cowan or Amari style neural field equations are used to form a parametric model, where the parameters are estimated from data. To illustrate the estimation framework, data is generated using the neural field equations incorporating modelled sensors, so a comparison can be made between estimated and true parameters. To facilitate state and parameter estimation, we introduce a method to reduce the continuum neural field model, using a basis function decomposition, forming a finite-dimensional state-space model. Spatial frequency analysis methods are introduced that systematically specify the basis function configuration required to capture the dominant characteristics of the neural field. The estimation procedure consists of a two-stage iterative algorithm incorporating the unscented Rauch-Tung-Striebel smoother for state estimation and a least squares algorithm for parameter estimation. The results show that it is theoretically possible to reconstruct the neural field and estimate intracortical connectivity and synaptic dynamics with the proposed framework. The results also illustrate a systematic loss of high spatial frequency information as the cost incurred by the model reduction procedure.

This framework provides a link between electrophysiological data and theoretical neural field models. This link may lead to greater understanding of cortical dynamics at the mesoscopic and macroscopic scales where diseases such as epilepsy are manifested.
% Please keep the Author Summary between 150 and 200 words
% Use first person. PLoS ONE authors please skip this step. 
% Author Summary not valid for PLoS ONE submissions.   
\section*{Author Summary}
With recent advances in medical imaging techniques and computing power, we are entering a new age of neuroscientific research, where patient-specific information and experimental data is becoming incredibly detailed. In order to make the best use of this information, data-driven mathematical models should be formed. This is particularly poignant where considering poorly understood complex systems such as the large-scale dynamics of the human brain. There currently exists a library of mathematical models describing large-scale brain dynamics, providing a theoretical explanation of interesting phenomena associated with various cognitive states and diseases. In this paper, we have created a framework for creating a data-driven mathematical model, forming a bridge between theory and electrophysiological data. Within this framework, we propose rules for designing experiments and we derive the mathematical formula to create the model. To demonstrate the utility of this framework, we provide an example using synthetic data that shows a comparison between true and estimated model characteristics. The ability to create patient-specific models of large-scale brain dynamics has the potential to revolutionise the treatment of diseases such as epilepsy, where techniques from engineering may be used to prevent abnormal neural oscillations.

\section*{Introduction}
Generating physiologically plausible neural field models is of great importance for studying brain dynamics at the mesoscopic and macroscopic scale. While our understanding of the function of neurons is well developed, the overall behaviour of the brain's mesoscopic and macroscopic scale dynamics remains largely theoretical. Understanding the brain at this level is extremely important since it is at this scale that pathologies such as epilepsy, Parkinson's disease and schizophrenia are manifested.

@@

For models to be clinically viable, they must be patient-specific. A possible approach to achieve this would be to fit a general continuum neural field model, like the Wilson and Cowan (WC)~\cite{Wilson1973} or Amari~\cite{Amari1977} models or a neural mass model like the Jansen and Ritt model~\cite{Jansen1995} to patient-specific data. Fitting the neural models to individuals is a highly non-trivial task and, until very recently, has not been reported in the literature.

An estimation framework for neural mass models known as dynamical causal modelling (DCM) \cite{David2003,David2006} has recently been proposed for studying evoked potential dynamics. Via a Bayesian inference scheme, DCM estimates the long range (cortico-cortical) connectivity structure between the specific isolated brain regions that best explains a given data set using the Jansen and Ritt equations~\cite{Jansen1995}. Another recent publication describing a parameter estimation method with a neural field model used an unscented Kalman filter with modified WC neural field equations~\cite{schiff2008kalman}. This work takes a system theoretic approach to the neural estimation problem, successfully demonstrating that it is possible to perform state estimation on spatiotemporal neural fields. This marks the first step in what has the potential to revolutionise the treatment of many neurological diseases where therapeutic electrical stimulation is viable.

We present an extension to the work of Schiff and Sauer~\cite{schiff2008kalman} by establishing a framework for estimating the state of the WC equations for larger scale (more space) systems via a systematic model reduction procedure. In addition, a method is presented for estimating the connectivity structure and the synaptic dynamics. Until now, model-based estimation of local intracortical connectivity has not been reported in the literature (to the best of the authors' knowledge). Our study builds on recent work which shows that it is possible to estimate local coupling of spatiotemporal systems using techniques from control systems theory and machine learning~\cite{Dewar2009}. The key development of this previous work was to represent a spatiotemporal system as a standard state-space model, with the number of states independent of the number of observations (recording electrodes in this case). In addition, the appropriate model selection tools have been developed~\cite{Scerri2009} allowing for the application of the technique to neural fields.

@@

Due to the relationship between the number of states in the model and the number of model parameters, increases in dimensionality lead to increases in uncertainty of the estimates. In addition, directly linking the complexity of the underlying system to the spatial resolution of the observation process implies that the system's dynamics become increasingly more complex the closer we look, which may not be an accurate modelling assumption.

The current study builds on this previous work~\cite{schiff2008kalman}, where the estimator is constructed to approximate a continuous field and the dimensionality of the state vector is independent of the resolution of the observation process and spatial resolution of the model. Instead, the dimensionality is governed by the expected complexity of the underlying field, which we measure using the frequency analysis presented in Methods section. These methods provide a rigorous framework for determining the dimensionality of the system using inherent spatial properties of the field. This enables estimation of a larger field with a lower dimensional state vector than would be otherwise possible.

The fact that the connectivity kernel is isotropic means that we are able to re-write the spatial mixing effect of the kernel with the firing rates as a convolution of the kernel with the membrane voltage field basis functions. This transformation greatly simplifies the computational complexity of the estimator, which is important for the estimator to run in a real-time scenario.

\subsection*{Patient-Specific Modelling}
This is the first paper to derive a continuum neural field model-based estimation framework for intracortical connectivity from electrophysiological data. The importance of the estimation framework can be illustrated by considering that the connectivity structure at this scale cannot be estimated using diffusion tensor imaging due to the near isotropic diffusion of water in grey matter~\cite{Assaf2008}.

@@

\subsection*{Future Work}

>>>>

% Please keep the abstract between 250 and 300 words
\section*{Abstract}

This paper presents a framework for creating neural field models from electrophysiological data. The Wilson and Cowan or Amari style neural field equations are used to form a parametric model, where the parameters are estimated from data. To illustrate the estimation framework, data is generated using the neural field equations incorporating modelled sensors, so a comparison can be made between estimated and true parameters. To facilitate state and parameter estimation, we introduce a method to reduce the continuum neural field model, using a basis function decomposition, to form a finite-dimensional state-space model. Spatial frequency analysis methods are introduced that systematically specify the basis function configuration required to capture the dominant characteristics of the neural field. The estimation procedure consists of a two-stage iterative algorithm incorporating the unscented Rauch-Tung-Striebel smoother for state estimation and a least squares algorithm for parameter estimation. The results show that it is theoretically possible to reconstruct the neural field and estimate intracortical connectivity and synaptic dynamics with the proposed framework. The results also illustrate a systematic loss of high spatial frequency information as the cost incurred by the model reduction procedure.

\section*{Introduction}
Generating physiologically plausible neural field models is of great importance for studying brain dynamics at the mesoscopic and macroscopic scales. While our understanding of the function of neurons is well developed, the overall behaviour of the brain's mesoscopic and macroscopic scale dynamics remains largely theoretical. Understanding the brain at this level is extremely important since it is at this scale that pathologies such as epilepsy, Parkinson's disease and schizophrenia are manifested.

@@

For models to be clinically viable, they must be patient-specific. A possible approach to achieve this would be to fit a general continuum neural field model, like the Wilson and Cowan (WC)~\cite{Wilson1973} or Amari~\cite{Amari1977} models, or a neural mass model like the Jansen and Ritt model~\cite{Jansen1995}, to patient-specific data. Fitting the neural models to individuals is a highly non-trivial task. Recently, however, this taks has been approached from a number of standpoints.

An estimation framework for neural mass models known as dynamical causal modelling (DCM) \cite{David2003,David2006} has been proposed for studying evoked potential dynamics. Via a Bayesian inference scheme, DCM estimates the long range (cortico-cortical) connectivity structure between the specific isolated brain regions that best explains a given data set using the Jansen and Ritt equations~\cite{Jansen1995}. Another recent approach estimates the parameters of a neural field model using an unscented Kalman filter based on modified WC neural field equations~\cite{schiff2008kalman}. This work takes a system theoretic approach to the neural estimation problem, successfully demonstrating that it is possible to perform state estimation on spatiotemporal neural fields. This marks the first step in what has the potential to revolutionise the treatment of many neurological diseases where therapeutic electrical stimulation is viable.

We present an extension to the work of Schiff and Sauer~\cite{schiff2008kalman} by establishing a framework for estimating the state of the WC equations for larger scale (more space) systems via a systematic model reduction procedure. In addition, a method is presented for estimating the connectivity structure and the synaptic dynamics. Until now, model-based estimation of local intracortical connectivity has not been reported in the literature (to the best of the authors' knowledge). Our study extends recent work which shows that it is possible to estimate local coupling of spatiotemporal systems using techniques from control systems theory and machine learning~\cite{Dewar2009}. The key development of this previous work was to represent a spatiotemporal system as a standard state-space model, with the number of states independent of the number of observations (recording electrodes in this case). In addition, the appropriate model selection tools have been developed~\cite{Scerri2009} allowing for the application of the technique to neural fields. This paper extends the linear framework of \cite{Dewar2009,Scerri2009} to the nonlinear case required by the WC model.

@@

This, in turn, will lead to increases in uncertainty of the estimates. In addition, directly linking the complexity of the underlying system to the spatial resolution of the observation process implies that the system's dynamics become increasingly more complex the closer we look, which may not be an accurate modelling assumption.

The current study builds on Schiff and Sauer's approach ~\cite{schiff2008kalman} by estimating a continuous field whose complexity is independent of the resolution of the observation process and spatial resolution of the model. Instead, the dimensionality is governed by the expected complexity of the underlying field, which we measure using the frequency analysis presented in Methods section. These methods provide a rigorous framework for determining the dimensionality of the system using inherent spatial properties of the field. This enables estimation of a larger field with a lower dimensional state vector than would be otherwise possible.

That the connectivity kernel is isotropic means the mixing effect of the kernel on the firing rates can be written as a convolution of the kernel with the membrane voltage field basis functions. This transformation greatly simplifies the computational complexity of the estimator, which is important for the estimator to run in a real-time scenario.

\section*{Future Work}

\subsection*{Patient-Specific Modelling}

@@

\subsection*{Extensions to the framework}
